Characterizing hippocampal replay using hybrid point process state space models

### 2019-11-23 paper.tex
Updates from Overleaf

\begin{abstract}
In the hippocampus, replay sequences are temporally compressed patterns of cell activity that resemble patterns that occur when the animal is moving through the environment. Because replay sequences typically occur when the animal is at rest, replay is hypothesized to be part of an internal cognitive process that enables the retrieval of past spatial memories and the planning of future movement. Traditionally, replay sequences have been discovered by looking for sharp wave ripples---high frequency oscillations that occur in association with replay---and then looking for spatially continuous patterns of cell activity; however, it has become clear that this does not fully depict the complexity of replay sequences. Replay sequences do not always co-occur with sharp wave ripples, have more complex dynamics than spatially continuous movement, have different temporal ordering than during movement, and change based on task. In this work, we introduce a hybrid state space framework to describe the richness of replay sequences. We show how defining discrete latent states associated with continuous latent dynamics and point process observations allows us to identify when replay sequences occur, categorize the type of sequence based on their proscribed continuous dynamics, and decode the spatial trajectory corresponding to the replay sequence.

@@

\subsection{Acausal Smoother}
Assuming the causal filter $p(x_{k}, I_{k} \mid \Delta N_{1:k}^{(1:C)})$ has been computed for all $k = 0, ..., T$, we want to compute $p(x_{k}, I_{k} \mid \Delta N_{1:T}^{(1:C)})$ for all $k < T$. This is known as the fixed-interval smoother and at any given time, it takes into account spiking observations in the past and future (except at 0 and time $T$). Similar to the filter posterior, this can also be expanded into a recursive formula that now uses the future smoother posterior $p(x_{k+1}, I_{k+1} \mid \Delta N_{1:T}^{(1:C)})$ and filter posterior $p(x_{k}, I_{k} \mid \Delta N_{1:k}^{(1:C)})$ at time step $k$.
\begin{multline}
p(x_{k}, I_{k} \mid \Delta N_{1:T}^{(1:C)}) = \\
p(x_{k}, I_{k} \mid \Delta N_{1:k}^{(1:C)}) \\
\times \sum_{I_{k+1}} \int \frac{p(x_{k+1} \mid x_{k}, I_{k+1}, I_{k}) Pr(I_{k+1} \mid I_{k})}{p(x_{k+1}, I_{k+1} \mid \Delta N_{1:k}^{(1:C)})} \\
\times p(x_{k+1}, I_{k+1} \mid \Delta N_{1:T}^{(1:C)}) dx_{k+1}
\end{multline}.

@@

\subsection{Estimation}
\section{Results}
\subsection{Example 1}
\begin{figure}[htbp]
\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}
\subsection{Example 2}
\section{Conclusion}
In this work, we have developed a point process state-space framework for understanding hippocampal replay that associates discrete latent states with continuous latent dynamics and allows for switching between the discrete states. We showed this framework is flexible: it can be deployed to characterize whether a replay represents spatially continuous trajectories or spatially discontinuous trajectories or it can be deployed to characterize local and non-local patterns of activity. Although we have only shown two examples of this framework, one could imagine combining the examples into one model---specifying that only during non-local states do we care about whether the trajectory is spatially continuous or not for example---or specifying more complicated conditions. All one has to specify is the initial condition, the likelihood, and state transition dynamic associated with each discrete state. The discrete state is then immediately interpretable in terms of the model.

This combination of interpretability and flexibility makes the framework is extremely powerful. By formulating what needs to be characterized mathematically, neuroscientists are forced to explicitly formulate their mental model of the replay characteristics of interest. This model can then be evaluated based on the data, criticized if the model assumptions are not adequate and improved into a better model. This iterative model building approach should be particularly powerful with respect to the hippocampus, where there are well-established features of the data and corresponding theories about the functioning of the hippocampus. The richness of data and theory allows neuroscientists to take advantage of both the flexibility of the framework and the specificity that it allows.

>>>>

\begin{abstract}
In the hippocampus, replay sequences are temporally compressed patterns of cell activity that resemble patterns that occur when the animal is moving through the environment. Because replay sequences typically occur when the animal is at rest, replay is hypothesized to be part of an internal cognitive process that enables the retrieval of past spatial memories and the planning of future movement. Traditionally, replay sequences have been discovered by looking for sharp wave ripples---high frequency oscillations that occur in association with replay---and then looking for spatially continuous patterns of cell activity; however, it has become clear that this does not fully depict the complexity of replay sequences. Replay sequences do not always co-occur with sharp wave ripples, have more complex dynamics than spatially continuous movement, have different temporal ordering than during movement, and change based on task. In this work, we introduce a hybrid state space framework to describe the richness of replay sequences. We show how defining discrete latent states associated with continuous latent dynamics and point process observations allows us to identify when non-local replay sequences occur, categorize the type of sequence based on their proscribed continuous dynamics, and decode the spatial trajectory corresponding to the replay sequence.

@@

\subsection{Acausal Smoother}
Assuming the causal filter $p(x_{k}, I_{k} \mid \Delta N_{1:k}^{(1:C)})$ has been computed for all $k = 0, ..., T$, we want to compute $p(x_{k}, I_{k} \mid \Delta N_{1:T}^{(1:C)})$ for all $k < T$. This is known as the fixed-interval smoother and at any given time, it takes into account spiking observations in the past and future (within the bounds of 0 and time $T$). Similar to the filter posterior, this can also be expanded into a recursive formula that now uses the future smoother posterior $p(x_{k+1}, I_{k+1} \mid \Delta N_{1:T}^{(1:C)})$ and filter posterior $p(x_{k}, I_{k} \mid \Delta N_{1:k}^{(1:C)})$ at time step $k$.
\begin{multline}
p(x_{k}, I_{k} \mid \Delta N_{1:T}^{(1:C)}) = \\
p(x_{k}, I_{k} \mid \Delta N_{1:k}^{(1:C)}) \\
\times \sum_{I_{k+1}} \int \frac{p(x_{k+1} \mid x_{k}, I_{k+1}, I_{k})
Pr(I_{k+1} \mid I_{k})}{p(x_{k+1}, I_{k+1} \mid \Delta N_{1:k}^{(1:C)})} \\
\times p(x_{k+1}, I_{k+1} \mid \Delta N_{1:T}^{(1:C)}) dx_{k+1}
\end{multline}.

@@

\subsection{Estimation}
\section{Results}
In order to demonstrate the flexibility of the framework, we show two brief examples of how associating different continuous state transitions and likelihoods with discrete states can lead to interpretable models which can distinguish between different features of replay.
\subsection{Continuous vs. Fragmented Spatial Trajectories}
The phenomenon of hippocampal replay is often conceived of as a spatially continuous trajectory through the environment, mimicking the animal's movements through the environment. However, it has been demonstrated that this is not always the case. For example, Karlsson and Frank \cite{KarlssonAwakereplayremote2009} showed that replays can sometimes correspond to the previously experienced environment -- which might appear as a spatially discontinuous trajectory--or what we will call here as fragmented--when decoding the replay relative to the current environment. Furthermore, since the hippocampus is also associated with episodic memory, it is possible that the replay event could correspond to some other non-spatial memory. Therefore, it is important to understand when the replay trajectory reflects the current environment versus something else. We can forumlate this 
\begin{figure}[ht]
\centerline{\includegraphics{fig1.pdf}}
\caption{The framework can be used to characterize changes between spatially continuous replay trajectories and spatially fragmented replay trajectories on simulated and real data. \textbf{(a)} The left plot shows a replay from 19 simulated neurons, arranged by the peak of their place fields, where the first three quarters of the replay is simulated to be spatially continuous and the last quarter is spatially fragmented. The black ticks indicate times when an action potential occured. The blue and orange lines indicate the estimated probability of the continuous and fragmented discrete states over time $Pr(I_{k} \mid \Delta N_{1:T}^{(1:C)})$, respectively. The right plot shows the posterior probability of position represented by the neurons over time, marginalized over the discrete states $p(x_{k} \mid \Delta N_{1:T}^{(1:C)})$. \textbf{(b)} An example of how the model can capture similar patterns in real data from the spiking patterns of 32 hippocampal neurons during a sharp wave ripple. Conventions are the same as in the first plot. Positions in the right plot correspond to linearized position of a W-shaped track, where 0 cm corresponds to the top of the middle tine of the W, 354 cm corresponds to the top of the left tine of the W, and 212 cm corresponds to the right tine of the W. The top of the center arm and bottoms of the left and right arm are connected.}
\label{fig1}
\end{figure}

\subsection{Local vs. Non-Local Coding}
\begin{figure}[ht]
\centerline{\includegraphics{fig2.pdf}}
\caption{An example of how the framework can be used to distinguish between local representations of position by the neurons---position that represents the animal's current location--and non-local trajectories on real data. The bottom plot shows spiking from 17 hippocampal neurons, arranged by linearized position and colored by the position that corresponds to the peak of their place field activity. The animal is at position 0. The top plot shows the estimated probability of the local and non-local discrete states over time $Pr(I_{k} \mid \Delta N_{1:T}^{(1:C)})$ in purple and pink respectively. The middle plot shows the posterior probability of position represented by the neurons over time, marginalized over the discrete states $p(x_{k} \mid \Delta N_{1:T}^{(1:C)})$. The white line in the middle plot indicates the animal's position. Position conventions are the same as in Fig.~\ref{fig1}b.}
\label{fig2}
\end{figure}
\section{Conclusion}
In this work, we have developed a point process state-space framework for understanding hippocampal replay that associates discrete latent states with continuous latent dynamics and allows for switching between the discrete states. We showed this framework is flexible: it can be deployed to characterize whether a replay represents spatially continuous trajectories or spatially discontinuous trajectories or it can be deployed to characterize local and non-local patterns of activity. Although we have only shown two examples of this framework, one could imagine combining the examples into one model---specifying that only during non-local states do we care about whether the trajectory is spatially continuous or not for example---or specifying more complicated conditions. All one has to specify is the initial condition, the likelihood, and state transition dynamic associated with each discrete state. The discrete state is then immediately interpretable in terms of the model. For example, the Poisson likelihood in the examples of this paper could easily be replaced by the marked point process likelihood in \cite{DengRapidclassificationhippocampal2016}---which would obviate the need for spike sorting before fitting the model.

This combination of interpretability and flexibility makes the framework is extremely powerful. By formulating what needs to be characterized mathematically, neuroscientists are forced to explicitly formulate their mental model of the replay characteristics of interest. This model can then be evaluated based on the data, criticized if the model assumptions are not adequate and improved into a better model. This iterative model building approach should be particularly powerful with respect to the hippocampus, where there are well-established features of the data and corresponding theories about the functioning of the hippocampus. The richness of data and theory about hippocampus could enable neuroscientists to take advantage of both the flexibility of the framework and the specificity that it allows.
